---
title: "Flight Data EDA"
output:
  html_document:
    df_print: paged
author: "Marie Bennett"
date: "2024-07-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Reading in Data
```{r}
flight <- read_csv("../Data_Files/Flights_Wrangled_Old.csv", col_names=TRUE)
head(flight)
```

# Selecting Possible Variables of Interest
```{r}
#selecting any variables that might be useful
flight_sub <- flight %>%
  select(Year, DayofMonth, DayOfWeek, Operating_Airline., Origin, OriginCityName, OriginState, OriginStateName, Dest, DestCityName, DestState, DestStateName, DepDelay, DepDelayMinutes, DepDel15, ArrDelay, ArrDelayMinutes, ArrDel15, Cancelled, Diverted, ActualElapsedTime, AirTime, Distance, DistanceGroup, CarrierDelay, WeatherDelay, NASDelay,SecurityDelay, LateAircraftDelay)
```

## Numeric Variables
```{r}
#numeric variable subset (not included: specific delays)
flight_num <- flight_sub %>%
  select(DepDelay, DepDelayMinutes, ArrDelay, ArrDelayMinutes, ActualElapsedTime, AirTime, Distance)
summary(flight_num)
```
```{r}
#histograms of each numeric variable
apply(X=flight_num, MARGIN=2, FUN=hist)
```
**Findings:**\
  - all numeric variables are right skewed\
  - might be able to transform ActualElapsedTime, AirTime, and Distance variable, but probably not the delay and arrival variables\
  - **idea**: using PCA might still work since normality is not required\
  
```{r}
#correlation matrix of numeric variables
cor(flight_num, use="complete.obs")
```

**Findings:**\
  - DepDelay, DepDelayMinutes, ArrDelay, and ArrDelayMinutes highly correlated (choosing to keep only DepDelay and ArrDelay)\
  - ActualElapsedTime, AirTime, and Distance highly correlated (choosing to keep AirTime and Distance)\
  - neither group is strongly correlated with the other
  - **idea**: could use PCA to break into a delay and a time component, this will get rid of correlation and help reduce the number of variables being investigated.\
  
```{r}
#delay type subset
flight_delay <- flight_sub %>%
  select(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay)
flight_delay <- na.omit(flight_delay)
head(flight_delay)
```
```{r}
summary(flight_delay)
```
**Findings:**\
  - skewed distributions\
  - clear high-end outliers\
  - **idea**: may not be able to use the minutes, but could create binary variables to indicate the type of delay\

## Categorical Variables
```{r}
#categorical variables subset
#filtered to only include flights between the top 10 airports
flight_cat <- flight_sub %>%
  select(Year, DayofMonth, DayOfWeek, Operating_Airline., Origin, OriginCityName, OriginState, OriginStateName, Dest, DestCityName, DestState, DestStateName, DepDel15, ArrDel15, Cancelled, Diverted, DistanceGroup) %>%
  filter((Dest == "ATL" | Dest == "DFW" | Dest == "DEN" | Dest == "ORD" | Dest == "LAX" | Dest == "CLT" | Dest == "MCO" | Dest == "LAS" | Dest == "PHX" | Dest == "MIA") & (Origin == "ATL" | Origin == "DFW" | Origin == "DEN" | Origin == "ORD" | Origin == "LAX" | Origin == "CLT" | Origin == "MCO" | Origin == "LAS" | Origin == "PHX" | Origin == "MIA"))

#frequency table of each variable
apply(X=flight_cat, MARGIN=2, FUN=table)
```
**Findings:**\
  - lose a lot of data with this filtering in the Operating_Airline variable for Alaska Airlines, Aligient Air, and maybe JetBlue\
  - DepDel15 and ArrDel15 do not have the same numbers with DepDel15 being slightly higher\
  - the number of cancelled and diverted flights is much smaller than not, but still enough to investigated differences\

```{r}
#possible predictors: DepDel15, ArrDel15, Cancelled, Diverted
#other variables: Year, DayofMonth, DayOfWeek, Operating_Airline., Origin, OriginCityName, OriginState, OriginStateName, Dest, DestCityName, DestState, DestStateName, DistanceGroup

#two-way frequency tables with Operating_Airline.
table(flight_cat$DepDel15, flight_cat$Operating_Airline.)
table(flight_cat$ArrDel15, flight_cat$Operating_Airline.)
table(flight_cat$Cancelled, flight_cat$Operating_Airline.)
table(flight_cat$Diverted, flight_cat$Operating_Airline.)

#two-way frequency tables with Origin
table(flight_cat$DepDel15, flight_cat$Origin)
table(flight_cat$ArrDel15, flight_cat$Origin)
table(flight_cat$Cancelled, flight_cat$Origin)
table(flight_cat$Diverted, flight_cat$Origin)

#two-way frequency tables with Dest
table(flight_cat$DepDel15, flight_cat$Dest)
table(flight_cat$ArrDel15, flight_cat$Dest)
table(flight_cat$Cancelled, flight_cat$Dest)
table(flight_cat$Diverted, flight_cat$Dest)

#two-way frequency tables with DistanceGroup
table(flight_cat$DepDel15, flight_cat$DistanceGroup)
table(flight_cat$ArrDel15, flight_cat$DistanceGroup)
table(flight_cat$Cancelled, flight_cat$DistanceGroup)
table(flight_cat$Diverted, flight_cat$DistanceGroup)
```
**Findings:**\
  - lack of data for certain airlines\
  - some origin airports might have more 15+ minute departure delays than others (proportionally)\
  - different origin airports generally have similar numbers in cancelled and diverted flights\
  - nothing of note for destination airports\
  - distance group 7 is very high in 15+ minute delays (maybe not, group 6 just has less values)\
  - possibly noting of note for distance groups

```{r}
# bar graph of days of the week split by cancellations
ggplot(flight, aes(x=DayOfWeek, fill=factor(Cancelled))) +
  geom_bar(position="dodge")

```
```{r}
# bar graph of days of the week split by 15 + minute delays
ggplot(flight, aes(x=DayOfWeek, fill=factor(DepDel15))) +
  geom_bar(position="dodge")
```

```{r}
# bar graph of quarters split by cancellations
ggplot(flight, aes(x=Quarter, fill=factor(Cancelled))) +
  geom_bar(position="dodge")
```
```{r}
# bar graph of quarters split by 15 + minute delays
ggplot(flight, aes(x=Quarter, fill=factor(DepDel15))) +
  geom_bar(position="dodge")
```
# July 9 Code
Two-way ANOVA with DepDelayMinutes as the response variable and Origin and DayOfWeek as the predictor variables.
```{r}
#exploration of the data
flight %>%
  group_by(Origin, DayOfWeek) %>%
  summarise(mean = mean(DepDelayMinutes, na.rm=TRUE),
            sd = sd(DepDelayMinutes, na.rm=TRUE),
            n = n())
```

```{r}
#checking assumptions
#independence of observations, equal variance of errors, normality of errors
lm_flight <- lm(DepDelayMinutes ~ factor(Origin) + factor(DayOfWeek), data = flight)
plot(lm_flight)
# constant variance might pass, but normality does not (is there a transformation for this?)
```

```{r}
#two-way ANOVA
origin_week_aov <- aov(DepDelayMinutes ~ factor(Origin) + factor(DayOfWeek), data = flight)
summary(origin_week_aov)
```
```{r}
ggplot(flight, aes(Origin, DepDelayMinutes, fill=factor(DayOfWeek))) +
  geom_col(position="dodge")
```

