---
title: "Investigating Taxi Out"
author: "Fred Lindsey"
date: "2024-07-18"
output: html_document
---


```{r include=FALSE}
library(tidyverse)
library(ggplot2)
```

```{r}
df = read.csv("../Data_Files/Flights_train.csv")
head(df)
```

```{r}
# Plot histogram of Departure Delay Minutes with logarithmic scale on y-axis
dep_dist <- ggplot(data=df, aes(x=DepDelayMinutes)) + geom_histogram(bins=100) + scale_y_log10()
dep_dist
```


```{r}
# Plot histogram of Taxi Out time
taxi_dist <- ggplot(data=df, aes(x=TaxiOut)) + geom_histogram(bins=100)
taxi_dist
```
Note: there's going to be a more explainable and interpretable signal in the data if outliers/influential points are removed from these two variables, TaxiOut and DepDelay. Start by looking at the % of data for each variable that exists beyond a certain point - 50 mins for TaxiOut, and 1500 mins for DepDelay. Then we can look at the scatterplot of those two data subsets for a more concrete relationship


```{r}
# Filter dataset to remove outliers from TaxiOut and DepDelayMinutes
delay_and_taxi_subset <- df %>% filter(TaxiOut < 51, DepDelayMinutes < 1501)

nrow(delay_and_taxi_subset)
nrow(df)
```


# Plot before manually removing outliers
```{r}
# Scatter plot of TaxiOut vs. DepDelayMinutes with color representing Airline
taxi_v_delay <- ggplot(data=df, aes(x=TaxiOut, y=DepDelayMinutes, color=factor(Airline))) + 
  geom_point() + 
  labs(title = "TaxiOut vs. DepDelay",
       x = "TaxiOut (minutes)",
       y = "Departure Delay (minutes)")

taxi_v_delay
```

# Plot after manually removing outliers
```{r}
# Scatter plot of TaxiOut vs. DepDelayMinutes for filtered data with color representing Airline
taxi_v_delay_v2 <- ggplot(data=delay_and_taxi_subset, aes(x=TaxiOut, y=DepDelayMinutes, color=factor(Airline))) + 
  geom_point() + 
  labs(title = "TaxiOut vs. DepDelay",
       x = "TaxiOut (minutes)",
       y = "Departure Delay (minutes)")

taxi_v_delay_v2
```

# Plot by airline using facet_wrap
```{r}
# Scatter plot of TaxiOut vs. DepDelayMinutes for filtered data, faceted by Airline
taxi_v_delay_facet <- ggplot(data=delay_and_taxi_subset, aes(x=TaxiOut, y=DepDelayMinutes)) + 
  geom_point() + 
  labs(title = "TaxiOut vs. DepDelay",
       x = "TaxiOut (minutes)",
       y = "Departure Delay (minutes)") + 
  facet_wrap(~Airline)

taxi_v_delay_facet
```

# Plot with linear model
```{r}
# Scatter plot with linear model fit line for TaxiOut vs. DepDelayMinutes, faceted by Airline
ggplot(data=delay_and_taxi_subset, aes(x = TaxiOut, y = DepDelayMinutes)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "TaxiOut vs. DepDelay",
       x = "TaxiOut (minutes)",
       y = "Departure Delay (minutes)") + 
  facet_wrap(~Airline)
```

# Plot with linear model without faceting
```{r}
# Scatter plot with linear model fit line for TaxiOut vs. DepDelayMinutes without faceting
ggplot(data=delay_and_taxi_subset, aes(x = TaxiOut, y = DepDelayMinutes)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "TaxiOut vs. DepDelay",
       x = "TaxiOut (minutes)",
       y = "Departure Delay (minutes)")
```

# Correlation analysis
```{r}
cor.test(delay_and_taxi_subset$TaxiOut, delay_and_taxi_subset$DepDelayMinutes)
```

```{r}
cor.test(delay_and_taxi_subset$DepDelayMinutes,delay_and_taxi_subset$TaxiOut)
```

```{r}
cor.test(df$TaxiOut, df$DepDelayMinutes)
```
**Takeaway: there is a very weak, negative correlation between TaxiOut Time, and DepDelayMinutes**

# Linear model analysis
```{r}
# Linear model of DepDelayMinutes as a function of Airline and TaxiOut interaction
temp_lm <- lm(DepDelayMinutes ~ factor(Airline)*TaxiOut, data=df)
plot(temp_lm)
```
